Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-03-08 docs/MEE_submission/manuscript_MEE.tex
final manuscript for submission to MEE

\clearpage
\section*{Abstract}
1. Research synthesis requires data extraction from primary studies with effect sizes for meta-analyses being calculated from summary statistics. However, exact values of such statistics are commonly hidden in figures.

2. Extracting summary statistics from figures can be a slow process that is not easily reproducible. Additionally, current software lacks an ability to incorporate important meta-data (e.g., sample sizes, treatment / variable names) about experiments and is not integrated with other software to streamline analysis pipelines.

3. Here we present the R package \pkg{metaDigitise} which extracts descriptive statistics such as means, standard deviations and correlations from the four plot types: 1) mean/error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. \pkg{metaDigitise} is user-friendly and easy to learn as it interactively guides the user through the data extraction process. Notably, it enables large-scale extraction by automatically loading image files, letting the user stop processing, edit and add to the resulting data fame at any point.

4. Digitised data can be easily re-plotted and checked, facilitating reproducible data extraction from plots with little inter-observer bias. We hope that by making the process of figure extraction more flexible and easy to conduct it will improve the transparency and quality of meta-analyses in the future.
%% 209 words

 Keywords: meta-analysis, comparative analysis, data extraction, \proglang{R}, reproducibility, figures, images, summary statistics

@@

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on processed data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or use in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations on figures, make corrections and access calibration and proceeded data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}A,B), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}C,D).

@@

\pkg{metaDigitise} recognises four main types of plot; Mean/error plots, box plots, scatter plots and histograms (\ref{fig:all_extract}). All plot types can be extracted in a single call of \fct{metaDigitise} and integrated into one output. Alternatively, users can process different plot types separately, using separate directories. All four plot types are extracted slightly differently (outlined below). Upon completing all images, or quitting, either summarised or processed data is returned (specified by the user through the \code{summary} argument). Summarised data consists of a mean, SD and sample size, for each identified group within the plot (should multiple groups exist). In the case of scatter plots, the correlation coefficient between x and y variables within each identified group is also returned. Processed data consists of a list with slots for each of the four figure types, containing the calibrated points that the user has clicked. This may be particularly useful in the case of scatter plots.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to `Import existing data', which returns previously digitised data, from sinlge or all figures. Alternately, the \fct{import\_metaDigitise} function returns the data of previous digitisations, but without user interaction, allowing easier integration into larger scripts. `Edit existing data' allows the user to re-plot or edit information or digitisations that have previously be done.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Line extraction is not particularly useful for most meta-analyses, although we recognise that it may be useful in other fields. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return processed rather than summary data) to estimate the parameters needed.

>>>>

\clearpage
\section*{Abstract}
\begin{enumerate} 
\item Research synthesis, such as meta-analysis requires the extraction of effect sizes from primary literature. Such effect sizes are calculated from summary statistics. However, exact values of such statistics are commonly hidden in figures.

\item Extracting summary statistics from figures can be a slow process that is not easily reproducible. Additionally, current software lacks an ability to incorporate important meta-data (e.g., sample sizes, treatment / variable names) about experiments and is not integrated with other software to streamline analysis pipelines.

\item Here we present the R package \pkg{metaDigitise} which extracts descriptive statistics such as means, standard deviations and correlations from the four plot types: 1) mean/error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. \pkg{metaDigitise} is user-friendly and easy to learn as it interactively guides the user through the data extraction process. Notably, it enables large-scale extraction by automatically loading image files, letting the user stop processing, edit and add to the resulting data fame at any point.

\item Digitised data can be easily re-plotted and checked, facilitating reproducible data extraction from plots with little inter-observer bias. We hope that by making the process of figure extraction more flexible and easy to conduct it will improve the transparency and quality of meta-analyses in the future.
\end{enumerate}

\vskip10pt
 \textbf{Keywords:} meta-analysis, comparative analysis, data extraction, \proglang{R}, reproducibility, figures, images, summary statistics

@@

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or use in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations on figures, make corrections and access calibration and proceeded data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. This is based on the calibration used in the \pkg{digitize} R package \citep{Poisot2011}. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}A,B), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}C,D).

@@

\pkg{metaDigitise} recognises four main types of plot; Mean/error plots, box plots, scatter plots and histograms (\ref{fig:all_extract}). All plot types can be extracted in a single call of \fct{metaDigitise} and integrated into one output. Alternatively, users can process different plot types separately, using separate directories. All four plot types are extracted slightly differently (outlined below). Upon completing all images, or quitting, either summarised or calibrated data is returned (specified by the user through the \code{summary} argument). Summarised data consists of a mean, SD and sample size, for each identified group within the plot (should multiple groups exist). In the case of scatter plots, the correlation coefficient between x and y variables within each identified group is also returned. Calibrated data consists of a list with slots for each of the four figure types, containing the calibrated points that the user has clicked. This may be particularly useful in the case of scatter plots.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to `Import existing data', which returns previously digitised data, from sinlge or all figures. Alternately, the \fct{getExtracted} function returns the data of previous digitisations, but without user interaction, allowing easier integration into larger scripts. `Edit existing data' allows the user to re-plot or edit information or digitisations that have previously be done.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Line extraction is not particularly useful for most meta-analyses, although we recognise that it may be useful in other fields. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

### 2018-03-08 docs/MEE_submission/supplements.tex
final manuscript for submission to MEE

\subsection{Obtaining Processed Data}

While \fct{metaDigitise} provides you with the summary statistics by default, for all plot types, in many cases the user may actually be interested in obtaining the processed digitised data from scatter plots (i.e. calibrated points). This is very easy to do by changing the default \code{summary} argument from TRUE to FALSE in \fct{metaDigitise}. Instead of providing you with summary statistics it will return a list containing four slots for each of the figure types (mean error, box plot, histogram and scatter plots). An example of a data object returned from digitising figures is as follows:

@@

Here, the user can easily access the list of processed scatter plot data by simply extracting the scatter plot slot:

@@

Additionally there is another function, \fct{import\_metaDigitise}, that imports all digitised data from a directory, as either summarised or processed data. For example,

\begin{CodeChunk}
\begin{CodeInput}
>R import_metaDigitise(".../FiguresToExtract", summary=TRUE)
\end{CodeInput}
{\scriptsize
\begin{CodeOutput}
                  filename    group_id         variable   mean  error error_type n    r    sd   plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
     002_Doe_2013_Fig1.png      setosa  Sepal length      5.00  0.11  se        50  NA    0.78  mean_error
     002_Doe_2013_Fig1.png  viriginica  Sepal length      6.59  0.18  se        50  NA    1.26  mean_error
     002_Doe_2013_Fig1.png  versicolor  Sepal length      5.94  0.14  se        50  NA    1.01  mean_error
     003_Doe_2013_Fig3.png      catana  Sepal length      4.95  0.36  sd        50  NA    0.36  histogram
\end{CodeOutput}
}
\end{CodeChunk}

>>>>

\subsection{Obtaining Calibrated Data}

While \fct{metaDigitise} provides you with the summary statistics by default, for all plot types, in many cases the user may actually be interested in obtaining the calibrated data from scatter plots (i.e. all calibrated points). This is very easy to do by changing the default \code{summary} argument from TRUE to FALSE in \fct{metaDigitise}. Instead of providing you with summary statistics it will return a list containing four slots for each of the figure types (mean error, box plot, histogram and scatter plots). An example of a data object returned from digitising figures is as follows:

@@

Here, the user can easily access the list of calibrated scatter plot data by simply extracting the scatter plot slot:

@@

Additionally there is another function, \fct{getExtracted}, that imports all digitised data from a directory, as either summarised or calibrated data. For example,

\begin{CodeChunk}
\begin{CodeInput}
>R getExtracted(".../FiguresToExtract", summary=TRUE)
\end{CodeInput}
{\scriptsize
\begin{CodeOutput}
                  filename    group_id         variable   mean  error error_type n    r    sd   plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
     002_Doe_2013_Fig1.png      setosa  Sepal length      5.00  0.11  se        50  NA    0.78  mean_error
     002_Doe_2013_Fig1.png  viriginica  Sepal length      6.59  0.18  se        50  NA    1.26  mean_error
     002_Doe_2013_Fig1.png  versicolor  Sepal length      5.94  0.14  se        50  NA    1.01  mean_error
     003_Doe_2013_Fig3.png      catana  Sepal length      4.95  0.36  sd        50  NA    0.36  histogram
\end{CodeOutput}
}
\end{CodeChunk}
